---
title: "Project.Rmd"
author: "Caggiano Paolo & Giardini Davide"
date: "2023-06-23"
output: pdf_document
---

Import the data:

```{r, include=FALSE}
library(nnet)
library(MASS)
library(ggplot2)
```

```{r}
df <- read.csv("Dataset/ProjectData.csv")
```



# Preliminary Operation to the Explanatory Variables

First of all, we are going to do some preliminary operation on the "instruction" variables. This is because the variable's levels are deeply unbalanced: class 7 ("No formal education") has only one observation, while class 6 ("Some primary school") has 25.
```{r}
table(df$instruction)
```
This is going to be a problem for the classification model. For this reason, we decide to aggregate classes 5,6,7 in one class, that is going to represent people with a primary instruction or lower.
```{r}
df$instruction[df$instruction == 6 | df$instruction == 7 ] <- 5
table(df$instruction)
```
Now, we are going to change the labels of the classes into 1-4 with 1 being the lower level of instruction and 4 being the highest. We are doing this because the regression function we are going to utilize later does not accept variables with missing levels (in this case 2).
```{r}
df$instruction[df$instruction == 1] <- 2
df$instruction <-  df$instruction - 1
df$instruction <-  5- df$instruction
table(df$instruction)
```

Now we are going to convert all the variables into ordered and unordered categorical variables.<br>
Instruction, Household members and Knowledge Score have to be converted into ordered categorical variables:

```{r}
df$instruction <- ordered(df$instruction, levels = c(1:4))
df$household_members <- ordered(df$household_members, levels = c(1:6))
df$know_score <- ordered(df$know_score, levels = c(0:7))
```

<br> Employment status and area are instead converted to unordered categorical variables:

```{r}
df$employment_status <- factor(df$employment_status, levels = c(1,2,4,5,6,9,10))
df$area <- factor(df$area, levels = c(1:5))
```

# Literacy Model

With this first model we want to understand which are the socio-economic factors that help to explain financial literacy among people. In other words, we are going to build a model that tries to explain the "Knowledge Score" that we built in the preprocessing phase.<br>
To do so, we are going to use a Proportional Odds Logistic Regression Model.

```{r}
modKnow <- polr(know_score ~ sex + area+ household_members + age + instruction + employment_status,
              data = df, Hess=TRUE)
step(modKnow)
```

The Akaike Information Criterion suggests that we should remove the variables related to area, employment status and household members. We therefore re-estimate the model:
```{r}
modKnow <- polr(formula = know_score ~ sex + age + instruction, data = df, 
    Hess = TRUE)
summary(modKnow)
```
Since the polr function does not automatically give us the p-value, we are going to compute them separately:
```{r}
summary_table <- coef(summary(modKnow))
pval <- pnorm(abs(summary_table[, "t value"]),lower.tail = FALSE)* 2
summary_table <- cbind(summary_table, "p value" = round(pval,5))
summary_table
```
Since the regression is logistic, we have to compute the exponential transformation to fully interpret them:

```{r}
exp(coef(summary(modKnow)))
```

# Retirement

With the next models, we are going to tackle questions related to retirement savings (QF8 and QF9).

## QF8 - Who does not think to have a good retirement plan?

```{r}
table(df$qf8)
```
There are 139 individuals that have not provided an answer for the question, we are going to create a subset that does not include these observations:

```{r}
dfR <- df[!(df$qf8==-99),]
dfR <- dfR[!(df$qf8==-97),]
dfR$qf8 <- ordered(dfR$qf8, levels = c(6:1))
```

Once again, we are going to use the "polr" function to estimate the Proportional Odds Logistic Regression Model:
```{r}
modRet1 <- polr(qf8 ~ sex + area+ household_members + age + instruction + employment_status +
                 know_score, data = dfR, Hess = TRUE)
step(modRet1)
```
We re-estimate the model utilizing only the variables that have been selected through the AIC criterion:
```{r}
modRet1 <- polr(formula = qf8 ~ area + age + instruction + employment_status, 
    data = dfR, Hess = TRUE)
summary(modRet1)
```
We compute the p-values:
```{r}
summary_table <- coef(summary(modRet1))
pval <- pnorm(abs(summary_table[, "t value"]),lower.tail = FALSE)* 2
summary_table <- cbind(summary_table, "p value" = round(pval,5))
summary_table
```
We compute the exponential transformation of the estimates for better interpretability:
```{r}
exp(coef(summary(modRet1)))
```

We are now going to estimate a regression only with knowledge score, in order to understand the association that this variable has with the answer qf8:

```{r}
modRet1 <- polr(qf8 ~ know_score, data = dfR, Hess = TRUE)
summary(modRet1)
```
Once again, the p-values are computed here:
```{r}
summary_table <- coef(summary(modRet1))
pval <- pnorm(abs(summary_table[, "t value"]),lower.tail = FALSE)* 2
summary_table <- cbind(summary_table, "p value" = round(pval,5))
summary_table
```


## QF9 - Who does utilize more secure tools for building their retirement fund?

We classify answer a-f and i as a stable/secure retirement plan (1), while all the other answer are considered unsecure (0).
We are interested in identifying those variables that are related to the choice of an unsecure retirement plan.

```{r, include=FALSE}
df <- read.csv("Dataset/ProjectData.csv")
df$instruction <- ordered(df$instruction, levels = c(7:1))
df$household_members <- ordered(df$household_members, levels = c(1:6))
df$know_score <- ordered(df$know_score, levels = c(0:7))
df$employment_status <- factor(df$employment_status, levels = c(1,2,4,5,6,9,10))
df$area <- factor(df$area, levels = c(1:5))
```


```{r}
# We create a new subset without the observation that have not given an answer for this question
dfR2 <- df[!(df$qf9_99==1),]
```

```{r}
# We create a new column that contain the sum of the columns related to secure retirement plans
dfR2$sum <- dfR2$qf9_1 + dfR2$qf9_2 + dfR2$qf9_3 +
  dfR2$qf9_4 + dfR2$qf9_5 + dfR2$qf9_6 + dfR2$qf9_9
# We transform the observation that have any value different from 0 in this new column to 1.
# In this way any observation that have at least one secure tool for building their
# retirement plan will be classified as 1.
# While all the other observation will remain equal to zero.
dfR2$sum[dfR2$sum != 0] <- 1
```

```{r}
dfR2$sum <- factor(dfR2$sum, levels = c(0,1))
```


We now estimate the model and apply the Akaike Information Criterion:
```{r}
modRet2 <- glm(sum ~ sex + area + household_members + age + instruction +
                 employment_status + know_score, data = dfR2, family = "binomial")
step(modRet2)
```

Now we re-estimate the model with the variables identified by the AIC:
```{r}
modRet2 <- glm(sum ~ sex + area + household_members + age + instruction
               + employment_status, data = dfR2, family = "binomial")
summary(modRet2)
```


We are now going to build a stacked bar-plot to further investigate the relationship between employment status and the answer to QF9:
```{r}
dfR2$employment_status <- as.integer(dfR2$employment_status)
dfR2$employment_status[dfR2$employment_status == 2] <- 1
dfR2$employment_status[dfR2$employment_status != 1] <- 0
library(ggplot2)
```
```{r}
# Stacked
ggplot(dfR2, aes(fill=factor(sum, levels=c(0,1)), y = after_stat(count), x=factor(employment_status, levels = c(0,1)))) + 
    geom_bar(position="stack", stat="count") +
    xlab("Employment Status") +
#    legend("topleft", legend = c("Unsecure tools for retirement", "Secure tools for retirement"))
    scale_fill_discrete(labels=c('Risky', 'Safe')) +
    guides(fill=guide_legend(title="Tools for Retirement:")) +
    scale_x_discrete(labels= c("Unemployed", "Employed"))
```


# Personal Finance

In this section we are going to tackle questions related to Personal Finance (savings): QF2, QF3, QF4, QF13

## QF3 - Who uses non-smart ways to save money?

We classify answer b, d, e as a secure way of saving money (1), while all the other answer are considered unsecure (0). We are interested in
identifying those variables that are related to the choice of an unsecure plan for personal savings.

```{r}
# We remove the observation that have not given an answer for this question (155)
dfPF2 <- df[!(df$qf3_99==1),]
```

```{r}
# The method is equal to the one in QF9
dfPF2$sum <- dfPF2$qf3_3 + dfPF2$qf3_6 + dfPF2$qf3_7
dfPF2$sum[dfPF2$sum != 0] <- 1
```

```{r}
dfPF2$sum <- factor(dfPF2$sum, levels = c(0,1))
```

```{r}
modPF2 <- glm(sum ~ sex + area + household_members + age + instruction +
                employment_status + know_score, data = dfPF2, family = "binomial")
step(modPF2)
```

```{r}
modPF2 <- glm(formula = sum ~ area + household_members + instruction +
                employment_status + know_score, family = "binomial", data = dfPF2)
summary(modPF2)
```

## QF4 - Who is not capable of sustaining an imporvise expense?

```{r}
# We remove the observation that have not given an answer for this question,
# and those who have not a personal income (78 + 255)
dfPF3 <- df[!(df$qf4 == -99),]
dfPF3 <- dfPF3[!(df$qf4 == -98),]
```

```{r}
# We transform all the observation that have not answered with 1 ("Yes") as 0 (negative category)
# This is because "not knowing" is considered a negative response to the question
dfPF3$qf4[dfPF3$qf4 != 1] <- 0
```

```{r}
dfPF3$qf4 <- factor(dfPF3$qf4, levels = c(0,1))
```

```{r}
modPF3 <- glm(qf4 ~ sex + area + household_members + age + instruction +
                employment_status + know_score, data = dfPF3, family = "binomial")
step(modPF3)
```

```{r}
modPF3 <- glm(qf4 ~ age + instruction + employment_status + know_score, 
              family = "binomial", data = dfPF3)
summary(modPF3)
```

We plot the distribution of knowledge score between people that responded positively and negatively to answer QF4, in order to further investigate the relationship between this two variables:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
dfPF3 <- dfPF3 %>% drop_na(know_score)
dfPF3_0 <- dfPF3[dfPF3['qf4'] == 0,]
dfPF3_1 <- dfPF3[dfPF3['qf4'] == 1,]
```

```{r}
require(gridExtra)
plot1 <- ggplot(dfPF3_0, aes(x=know_score)) +
    geom_histogram(binwidth=.5, colour="black", fill="light green", stat="count") +
    ggtitle("Negative Answer (0)")  +
    theme(plot.title = element_text(hjust = 0.5)) +
    ylim(c(0,200))
plot2 <- ggplot(dfPF3_1, aes(x=know_score)) +
    geom_histogram(binwidth=.5, colour="black", fill="light blue", stat="count") +
    ggtitle("Positive Answer (1)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    ylim(c(0,200))
grid.arrange(plot1, plot2, ncol=2)
```




## QF13 - Who does not have an emergency fund?

This question evaluates if the person have an "Emergency Fund". According to the popular opinion, an emergency fund should cover at least 3-6 months of expenses. For this reason, answer d and e are considered positive (1), while all the other are considered negative (0).

```{r}
# We remove the observation that have not given an answer for this question
dfPF4 <- df[!(df$qf13 == -99),]
```

```{r}
dfPF4$qf13[dfPF4$qf13 == 1 | dfPF4$qf13 == 2 | dfPF4$qf13 == 3 | dfPF4$qf13 == -97] <- 0
dfPF4$qf13[dfPF4$qf13 != 0] <- 1
```

```{r}
dfPF4$qf13 <- factor(dfPF4$qf13, levels = c(0,1))
```

```{r}
modPF4 <- glm(qf13 ~ sex + area + household_members + age + instruction
              + employment_status + know_score, data = dfPF4, family = "binomial")
step(modPF4)
```

```{r}
modPF4 <- glm(formula = qf13 ~ area + age + employment_status + know_score, 
    family = "binomial", data = dfPF4)
summary(modPF4)
```



